---
title: "In-class 2/14/19"
author: "S670"
date: "2/14/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE)
library(tidyverse)
library(lattice)
cb_palette = c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

## Contour plots

**READ: Cleveland pp. 228--248.**

### Velocities of the NGC 7531 galaxy

The data frame `galaxy` contains 323 measurements of velocity for locations in galaxy NGC 7531.

```{r}
load("lattice.RData")
summary(galaxy)
```

`velocity` (in km/second) is the response variables, and the locations along `east.west` and `north.south` coordinates are the main explanatory variables. (North and west are positive -- if you lie on your back with your head pointing north, then west is to your right.) First, we simply look where the measurements were taken:

```{r}
ggplot(galaxy, aes(x=east.west, y=north.south)) + geom_point() + coord_fixed()
```

We see the measurements were taken along seven different lines ("slits".) The variables `radial.position` and `angle` give the locations in polar coordinates.

How does velocity vary with location? We first facet by the `velocity` variable.

```{r}
ggplot(galaxy, aes(x=east.west, y=north.south)) + geom_point() + geom_jitter(width=0.5, height=0.5) + facet_wrap(~cut_number(velocity, n=12), ncol=6) + coord_fixed() + labs(title="Galaxy locations split by velocity")
```

The slowest locations are in the top left panel -- these are all in the northeast. The fastest locations in the bottom right panel -- these are in the southwest.

We can also look at the velocities as a function of `radial.position`, faceted by `angle`.

```{r, message=FALSE}
ggplot(galaxy, aes(x=radial.position, y=velocity)) + geom_point() + geom_smooth() + facet_wrap(~angle, ncol=4) + labs(title="Galaxy position and velocity split by slit angle")
```

For each angle, we see a nonmonotonic relationship. The magnitude of the variation in velocities varies a lot -- it's big at 92.5 to 111 degrees (NNE to SSW), and small for 12.5 and 43 degrees (WNW to ESE.)

### Modeling galaxy velocities

We fit a full loess model with interaction. We add the argument `normalize=FALSE`, which is appropriate when the two explanatory variables are on the same scale and we don't want to standardize them. By trial and error, a smoothing parameter `span = 0.25` seems about right.

```{r, message=FALSE}
galaxy.lo = loess(velocity ~ east.west * north.south, data = galaxy, span = 0.25, family="symmetric", normalize=FALSE)
```

We create a data frame with the original variables as well as the fitted values and residuals:


```{r}
galaxy.lo.df = data.frame(galaxy,
  .fitted = fitted.values(galaxy.lo),
  .resid = residuals(galaxy.lo))
```

Now redraw the above plot faceted by angle with the fit added.

```{r}
ggplot(galaxy.lo.df, aes(x=radial.position, y=velocity)) + geom_point() + geom_line(aes(x=radial.position, y=.fitted), color="blue") + facet_wrap(~angle, ncol=4) + labs(title="Galaxy position and velocity split by slit angle with fitted model")
```

The model provides the right general shape to the data in polar coordinates. We take a closer look at the residuals:

```{r, message=FALSE}
gg = ggplot(galaxy.lo.df, aes(x=radial.position, y=.resid)) + geom_point() + geom_smooth(span=1) + facet_wrap(~angle, ncol=4)
gg + labs(title="Galaxy position and residuals split by slit angle with fitted model")
```

The 92.5 degrees plot is somewhat bad -- it slopes down and it appears heteroskedastic. There's a bit more slope and/or curvature in the others than we would expect under random error as well. Nevertheless, we'll keep our model for simplicity.

We now do our usual residual diagnostics. The spread-location plot of transformed residuals:

```{r, message=FALSE}
ggplot(galaxy.lo.df, aes(x=.fitted, y=sqrt(abs(.resid)))) + geom_point() + geom_smooth(method.args=list(degree=1))
```

The magnitude of the residuals may increase slightly with fitted value, but it's not so bad that it would demand a transformation. It is so bad that we can't do probabilistic stuff, so we omit the normal QQ plot.

Finally, how much of the variance does our model capture?

```{r}
var(galaxy.lo.df$.fitted) / var(galaxy$velocity)
```

Our model captures 104% of the of the variance of the galaxy velocities. Wait, what?

With complex models, the variance of the fitted values can sometimes exceed the variance of the response. If this bothers you, an alternative way of defining a (pseudo) $R^2$ is

$$
\frac{\textrm{variance of fitted values}}{\textrm{variance of fitted values} + \textrm{variance of residuals}}
$$

Under this definition, our $R^2$ becomes

```{r}
var(galaxy.lo.df$.fitted) / (var(galaxy.lo.df$.fitted) + var(galaxy.lo.df$.resid))
```

Some authors prefer using the squared correlation between the fitted values and the response:

```{r}
cor(galaxy.lo.df$.fitted, galaxy$velocity)^2
```

The difference is academic, though. Our model captures the vast majority of the variation in the galaxy velocities.

### Visualizing the fit: geom_raster and geom_contour

To prepare for plotting, let's make predictions on a rectangular grid, and put the results in a data frame, `galaxy.plot.df`.

```{r}
galaxy.grid = expand.grid(east.west=seq(-25,25,0.5), north.south=seq(-45,45,0.5))
galaxy.predict = predict(galaxy.lo, newdata=galaxy.grid)
galaxy.plot.df = data.frame(galaxy.grid, fit=as.vector(galaxy.predict))
```

Because of the interaction in the loess, the techinques we've used previously that condition on explanatories aren't going to be sufficient. One option is to use `geom_raster()` to indicate the third variable using color. (`geom_tile()` can be used instead for non-rectangular data.)

```{r}
ggplot(galaxy.plot.df, aes(x=east.west, y=north.south, fill=fit)) + geom_raster() + coord_fixed() + scale_fill_distiller(palette="RdYlBu")
```

It's often quite hard to find a color palette that clearly displays all the features you want to make apparently.
One solution is to facet on the fitted value. We cut the above plot into a large number of pieces by fitted value, then draw a grid of plots.

```{r}
ggplot(galaxy.plot.df, aes(x=east.west, y=north.south, fill=fit)) + geom_raster() + coord_fixed() + scale_fill_distiller(palette="RdYlBu") + facet_wrap(~cut_number(fit, n=16), ncol=4)
```

The basic pattern is clear: the fitted velocities are small in the northeast and generally increase as you move toward the southwest, though the details are complicated.

Alternatively, we can make the shapes clear on the original graph by adding **contours**. A contour is a curve that joins together points that have the same value of the $z$-variable. `geom_contour()` gives the basic contour plot:

```{r}
ggplot(galaxy.plot.df, aes(x=east.west, y=north.south, z=fit)) + geom_raster(aes(fill = fit)) + coord_fixed() + scale_fill_distiller(palette="RdYlBu") + geom_contour()
```

We can see, for example, a circle of points in the southwest all have the same fitted value. The coloring makes it clear this is a high value (velocity over 1700 km/second.) The northeast has the lowest fitted values.

A final choice is to skip the raster part and go straight to the contours.

```{r}
ggplot(data.frame(galaxy.grid, fit=as.vector(galaxy.predict)), aes(x=east.west, y=north.south, z=fit)) + geom_contour(binwidth=10, aes(color=..level..)) + coord_fixed()
```

If relying on contours alone, they need to be fairly dense to get the details of the distribution across. Note that labeling the contours by putting numbers directly on the graph is very hard to do within `ggplot` (if you must, use `contourplot` in the `lattice` library instead.)

### Contour plots for bivariate densities

The other major use for contour plots is to display density estimates for bivariate data. The ggplot function `stat_density_2d` does bivariate density estimation. We apply it to the data set `geyser`, which contains bivariate data on the waiting time until eruption and the duration of the subsequent eruption (both in minutes) for 299 successive eruptions of Old Faithful in 1985:

```{r}
library(MASS)
ggplot(geyser, aes(x=waiting, y=duration)) + geom_point() + stat_density_2d()
```

The data appears trimodal: we get short wait/long eruption, long wait/long eruption, and long wait/short eruption. We don't get short wait/short eruption.

We could also look at this by coloring in between the contours to get a solid surface:

```{r}
ggplot(geyser, aes(x=waiting, y=duration)) + stat_density_2d(aes(fill=..level..), geom="polygon")
```

The general message is the same, but now we have quantitative information on the density. The highest peak is for long wait/long eruption, at around 0.015 (per minute per minute.) You can explore the 2D density further by studying things like the conditional distribution of eruption length given waiting time, etc.

If you're sitting around at Yellowstone, you might be more interested in the joint density of the eruption duration and the waiting time to the *next* eruption.

```{r}
n = nrow(geyser)
next.geyser = data.frame(duration = geyser$duration[-n], waiting = geyser$waiting[-1])
ggplot(next.geyser, aes(x=duration, y=waiting)) + stat_density_2d(aes(fill=..level..), geom="polygon")
```

This time the data looks bimodal: short eruption followed by short wait, or long eruption followed by long wait. The long/long peak is substantially higher.

Finally, we want to study the **conditional** distributions: given an eruption duration, how long will we have to wait for the next eruption? For simplicity, we only split `duration` into two categories.

```{r}
duration.cat = rep(NA, nrow(next.geyser))
duration.cat[next.geyser$duration<=3] = "Short eruption"
duration.cat[next.geyser$duration>3] = "Long eruption"
duration.cat = factor(duration.cat, levels=c("Short eruption", "Long eruption"))
ggplot(data.frame(next.geyser, duration.cat), aes(x=waiting, fill=duration.cat, color=duration.cat)) + stat_density() + facet_wrap(~duration.cat, ncol=1) + scale_color_manual(values = cb_palette) + scale_fill_manual(values = cb_palette)
```

The modal waiting time after a short eruption is a bit over 50 minutes, while the modal waiting time after a long eruption is a bit under 80 minutes. Note that this graph does not attempt to show that long eruptions are more common than short eruptions.

## Truly 3D plots: Wireframes

**READ: Cleveland pp. 249--267.**

It's the lecturer's opinion that 3D plots are kind of overrated, but some people like them so here they are. `ggplot()` doesn't really do 3D plots, so we go to the `lattice` library for this section.

### Galaxy data

Let's return to the loess fit to the galaxy data from last time. Recall that north and west are positive; putting north on top, the slow parts of the galaxy were in the top left (northeast) and the fast parts of the galaxy were in the bottom right (southwest.) Firstly, we can use the `cloud()` function to plot the raw data in three dimensions.

```{r, message=FALSE}
load("lattice.RData")
cloud(velocity ~ east.west * north.south, data=galaxy)
```

The arrows point toward more positive values of the variables: north, west, and high velocity respectively. (If you wanted numeric scales on the axes instead of just arrows, you could use `scatterplot3d()` in the library of the same name, but I find it near-impossible to accurately read numbers off 3D plots so I don't bother including them.) Now that we know what to look for, we see the data is consistent with low velocities in the northeast and high velocities in the southwest. This would be possible but somewhat hard to see if we didn't know what we were looking for.

Now re-fit the loess model we chose last time, and make predictions on a grid.

```{r}
galaxy.lo = loess(velocity ~ east.west * north.south, data = galaxy, span = 0.25, family="symmetric", normalize=FALSE)
galaxy.wf.grid = expand.grid(east.west=seq(-25,25,2), north.south=seq(-45,45,2))
galaxy.wf.predict = predict(galaxy.lo, newdata=galaxy.wf.grid)
galaxy.wf.df = data.frame(galaxy.wf.grid, fit=as.vector(galaxy.wf.predict))
```

To draw a truly 3D plot, we use `wireframe()`.

```{r}
wireframe(fit ~ east.west * north.south, data=galaxy.wf.df)
```

This is quite a bit clearer than the `cloud()` plot. In particular, the curvature of the fitted surface is apparent. The estimated velocity goes down a bit in the extreme southwest corner.

With a wireframe plot, it's rare that one angle lets you see all relevant details of the fit. If you're just exploring the data on your own, the `rotate.wireframe()` function in the `TeachingDemos` library can be fun to play with (if a bit buggy.)

```{r, eval=FALSE}
# install.packages("TeachingDemos")
library(TeachingDemos)
rotate.wireframe(fit ~ east.west * north.south, data=galaxy.wf.df)
```

Of course, this doesn't fly if you're trying to prepare a document. You could build a Shiny app but that probably isn't going to be worth the effort. Instead, pick a few different angles, and show the wireframe from those angles.

```{r}
wireframe(fit ~ east.west * north.south, data=galaxy.wf.df, screen = list(z=30, x=-60, y=0))
wireframe(fit ~ east.west * north.south, data=galaxy.wf.df, screen = list(z=120, x=-60, y=0))
wireframe(fit ~ east.west * north.south, data=galaxy.wf.df, screen = list(z=210, x=-60, y=0))
wireframe(fit ~ east.west * north.south, data=galaxy.wf.df, screen = list(z=300, x=-60, y=0))
```

Here we kept the $x$ and $y$ angles fixed, while rotating $z$ by 90 degrees each time. The effect is to "spin" the surface around while keeping the "camera" fixed.

Finally, we can "fill in" in the wireframe using the `drape` argument.

```{r}
wireframe(fit ~ east.west * north.south, data = galaxy.wf.df, screen = list(z=120, x=-60, y=0), drape=TRUE)
```

You can play around with the color scheme using the `col.regions` argument. All this is getting dangerously close to chartjunk territory, though.

### Ethanol data

We now return to the ethanol engine data. Recall that the curves relating NOX to equivalence ratio (E) were close in shape but not identical for all five values of compression ratio (C). If we cloud plot the data and look at it "front-on," i.e. with C going into the screen, we get something that's almost a 2D scatterplot of NOx against E.

```{r}
cloud(NOx ~ C * E, data=ethanol, screen = list(z=90, x=-90, y=0))
```

We can try coloring the plot by levels of C:

```{r}
cloud(NOx ~ C * E, data=ethanol, screen = list(z=90, x=-90, y=0), groups=C)
```

Now we can see the blue points (for example) are lower than the others on the right hand side, while they're similar to the other colors on the left. (Again, this is easier to see in retrospect after having studied the data.) We could try to make this plot look nicer by fiddling with the color scheme or adding a legend, but we'll be better off instead looking at the loess surface we fitted to the data. Re-fit the model and look at it head-on:

```{r}
ethanol.lo = loess(NOx ~ C * E, data=ethanol, span=1/3, parametric="C", drop.square="C", family="symmetric")
ethanol.grid = expand.grid(C=c(7.5,9,12,15,18), E=seq(0.6, 1.2, 0.1))
ethanol.predict = predict(ethanol.lo, newdata=ethanol.grid)
ethanol.df = data.frame(ethanol.grid, fit=as.vector(ethanol.predict))
wireframe(fit ~ C * E, data=ethanol.df, screen=list(z=90, x=-90, y=0))
```

We see that the height of the surface (the fitted value) gets taller as C increases. Note, however, that changing the angles by just a few degrees make this seem to disappear.

```{r}
wireframe(fit ~ C * E, data=ethanol.df, screen=list(z=92, x=-97, y=0))
```

The issue is *foreshortening* -- things that are further away look smaller. To better understand foreshortening, look at a bunch of Italian Renaissance paintings. For our purposes, it's enough to remember that the choice of angle is important, so make sure to try out a few.

For this fit, looking somewhat from the side makes the change in height clearer:

```{r}
wireframe(fit ~ C * E, data=ethanol.df, screen=list(z=30, x=-60, y=0))
```

Now it's obvious that higher C generally means somewhat higher NOx. Again, you need more than one angle to see all of what's going on.

### Soil

The Cleveland data set `soil` contains measurements on resistivity (in ohm cm) in a field in Western Australia. The locations are given by "easting" and "northing" coordinates, which just measure distance from an origin in kilometers.

```{r}
ggplot(soil, aes(x=easting, y=northing)) + geom_point() + coord_fixed()
```

We see the locations occur along a number of "tracks", which are recorded in the `track` variable. Some tracks are north-south while others are east-west; this is recorded in the variable `is.ns`. To see how the resistivity varies by coordinate, we'll first subset by track direction, then facet by track.

```{r}
ggplot(subset(soil, is.ns == TRUE), aes(x=northing, y=resistivity)) + geom_point() + facet_wrap(~track, ncol=4)
```

Looking at the north-south tracks, the patterns aren't very consistent. There are a bunch of spikes occurring at seemingly random locations.

Now try the east-west tracks:

```{r}
ggplot(subset(soil, is.ns == FALSE), aes(x=easting, y=resistivity)) + geom_point(size=0.5) + facet_wrap(~track, ncol=8)
```

We see that there's usually a downward trend in resistivity as the easting coordinate increases. However, this doesn't show up in all the plots.

We fit a loess model to the data, predicting resistivity from the easting and northing coordinates (with an interaction.)

```{r}
soil.lo = loess(resistivity ~ easting * northing, span = 0.25, data=soil)
```

Now predict on a grid and plot the fit using color and contours.

```{r}
soil.grid = expand.grid(easting = seq(0, 1.5, 0.01), northing = seq(0, 3.5, 0.01))
soil.predict = predict(soil.lo, newdata=soil.grid)
soil.df = data.frame(soil.grid, fit=as.vector(soil.predict))
ggplot(soil.df, aes(x=easting, y=northing, z=fit, fill=fit)) + geom_raster() + geom_contour(binwidth=10, color="black") + scale_fill_distiller(palette="RdBu") + coord_fixed()
```

There's a clear peak around $(0.75, 2.1)$, along with a smaller peak near $(0.6, 0.9)$.

The complexity of the surface means that a wireframe plot isn't well-suited to displaying the fit. You can use `surf3D()` in the `plot3D` library:

```{r}
# install.packages("plot3D")
library(plot3D)
east.grid = seq(0, 1.5, 0.01)
north.grid = seq(0, 3.5, 0.01)
mesh.grid = mesh(east.grid, north.grid)
fit.grid = matrix(soil.predict, nrow=length(east.grid))
surf3D(mesh.grid$x, mesh.grid$y, fit.grid, theta=0, col = ramp.col(col = c("blue", "red"), n = 10))
```

It's unclear whether this is of any real data analytic value. It looks cool, I guess!







