---
title: "In-class 1/29/19"
author: "S670"
date: "1/29/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE, message = FALSE)
```

```{r}
library(tidyverse)
cb_palette = c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

Load GDP data:

```{r}
gdp = read.csv("GDPC1.csv")
```

Create a time series object:

```{r}
gdp.ts = ts(data = gdp$GDPC1, start = 1947, frequency = 4)
library(forecast)
autoplot(gdp.ts) + scale_y_log10() + geom_smooth()
```

Check to see if seasonal component has been removed.

Use `decompose()` to fit a fixed seasonal component:

```{r}
gdp.add = decompose(gdp.ts)
autoplot(gdp.add)
```

Try a multiplicative effect:

```{r}
gdp.multi = decompose(gdp.ts, type = "multiplicative")
autoplot(gdp.multi)
```

Using `stl()` lets you fit a time-varying seasonal effect, with the degree of smoothing determined by `s.window` (among other things), which should be a odd number.

```{r}
gdp.stl = stl(log(gdp.ts), s.window = 49, s.degree = 1)
autoplot(gdp.stl)
```

Forget about seasonal effects. Take out the trend:

```{r}
gdp.df = data.frame(realGDP = as.matrix(gdp.ts), year = time(gdp.ts))
gdp.lo = loess(log(realGDP) ~ year, data = gdp.df)
library(broom)
gdp.lo.df = augment(gdp.lo)
ggplot(gdp.lo.df, aes(x = year, y = .resid)) + geom_line(alpha = 0.5) + geom_smooth(method = "loess", span = .1, se = FALSE)
```












